When the Pen Becomes a Sword: Retired US Military Officers and Op-Ed Use in Contemporary America

### 2020-08-28 code/Submissionv3.Rmd
commit

# When Words Become Weaponized - A Theory of Credibility and Polarization

Figure 1, below, shows describes a theory for how the variables of political polarization, institutional credibility, military professionalism, and other military outcomes may relate to each other. From left to right, the theory contends that increasing political polarization ultimately reduces military professionalism. At the same time, however, there is an intervening variable, institutional credibility, that may actually increase as a result of increasing political polarization. Finally, on the far right of the diagram, are changes in military outcomes. Though it is outside of the scope of this paper, I contend that eventually, increases in political polarization may at some point cause discernible changes in various military outcomes. These military outcomes may vary, but could include things such as military effectiveness, to include the military's ability to fight and win battles [@talmadge_dictators_2015] and to formulate sound strategic advice [@rapp_civil-military_2015]. Indeed, the full degree to which and in what ways polarization will impact the military are not yet fully known.

The theory grounds the two independent variables discussed here, credibility and polarization, as changes that have occurred within American society in the past fifty years. Because these changes stem from changes that have their originate within broader American society, my theory contends that the military as an institution and the military professionals that comprise it are not impervious to their impacts. In other words, my theory assumes that maintaining professional conduct by the military institution requires a deliberate effort and is hard work, despite the fact that behaving in this manner is a long and storied norm. Professional conduct is not a guaranteed nor automatic outcome.

When institutional credibility is high, a demand signal is created by the American public such that future interactions between the military institution and the American public are encouraged and welcomed. From an institutional perspective, the military might respond to the public's demand signal for increased military interaction by increasingly participating in public ceremonies, parades, and other showcases of military prowess (at sporting venues, for instance). From an individual perspective, military elites, especially retired military officers, would be psychologically primed to increasingly dialogue with the American public when military credibility is high. By extension, other outlets such as the press would also be primed to showcase military elites when military credibility runs high, because these elites represent an institution that commands a high degree of public respect.

@@

problem_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MO"="Changes in\n Military\n Outcomes", 
                     "MP"="Reduction in\n Military\n Professionalism", 
                     "PP"="Increasing\n Political\n Polarization", 
                     "IC"="Rising\n Institutional\n Credibility")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure 1", subtitle="The Impact of Political Polarization on Military Professionalism")

@@

The degree and character of political polarization in a society also has the potential to impact the professionalism of the military as an institution. To understand how, recall that the three salient characteristics of the political polarization prevalent in contemporary America include an increase in the number and scope of issues that have become polarized (the phenomenon referred to as "conflict extension" [@layman_party_2002]), an increase in the strength of negative feelings held by supporters of an issue toward the opposition (the rise of "negative partisanship" [@abramowitz_great_2018]), and an increase in the salience of core issues related to individual and national identity [@white_values_2002; @mann_its_2012; @sides_identity_2018]. These three characteristics convey several implications for the military as an institution and its leaders.

First, there is increasing difficulty for individuals, to include military professionals, to remain neutral as polarized politics subsumes more and more issues, many of which involve issues which individuals hold to dearly. In previous eras, when polarization was not as extreme as it is today, it was more possible to stay out of politics. Today, polarization seems to do the opposite: polarized politics seem to thrust issues and the unfolding political fight at the individual, even when the individual is not seeking confrontation. A second implication of polarization is that it raises the salience for individuals of right and wrong in a moral sense, which increases the odds of an individual expressing partisan sentiment. In other words, in saying or doing what one believes to be right, one is increasingly likely to also be taking a partisan position. The third and most important implication for the military professional as a result of contemporary political polarization is that it impacts and distorts the notions of two character traits that are held dearly by members of the military profession: loyalty and duty.

As any soldier knows, few traits are more sacred in the military than loyalty and duty. These virtues are extolled in a host of codes and mantras that have played a central part of the American military since its inception. They are captured in phrases that military professionals say to themselves and to each other: "do the right thing," "lead with integrity," and, as the West Point Cadet Prayer so elegantly states, choose "the harder right over the easier wrong."

Political polarization, however, obfuscates these ideals, and places greater responsibility for interpreting the meaning and the associated action of these ideals on the individual and less on the corporate military institution. This stems from the fact that highly polarized political climate may involve specific issue areas that certain members of the military feel a stronger affinity towards than other members. Two different military officers may adopt completely different opinions of a contentious issue, each believing that the other's opinion is wrong. Should the officer act on that opinion, the other might believe he or she is operating outside of the proper military frameworks, or outside of what the right thing to do is according to the Constitution or other loyalty structures. Each officer, then, might accuse the other of failing to be loyal to the ideals of the military.

@@

In summary, it can be said that high degrees of political polarization impact military professionalism in key ways. Polarization increases the chances of expressing partisan sentiment in the name of doing what is morally right. Polarization, when it is high, also increases the chances of speaking out on issue areas that are outside of the realm of traditional military expertise, especially as the degree of polarization and "conflict extension" [@layman_party_2002] spreads, which reduces the perceived non-partisan space available to military professionals. Finally, polarization muddies the concepts of loyalty and duty. Strong feelings about issue areas perceived as crucial and essential by some greatly increase the odds of charging others of failing to act loyally or to fulfill one's duty.

Ultimately, four possible worlds exist when the impacts of military credibility and political polarization are considered in tandem (see Table 1, below). The first is when political polarization and military credibility are both low. When these conditions hold, there is no anticipated drop in military professionalism by the military institution nor its individual elites. With these conditions, the military does not have an out-sized platform to interact with the American public, nor are there outstanding incentives for the military to raise significant disagreement over any domestic political issues.

\singlespacing
\begin{table}[ht]
\begin{footnotesize}
\caption{Four Types of "Military Professionalism Worlds" Where Institutional Credibility and Political Polarization Vary} % title of Table
\centering % used for centering table
\begin{tabular}{|p{5cm}|p{5cm}|p{5cm}|p{5cm}|}
    \hline
    Influence of Variables & \textbf{Low Institutional Credibility} & \textbf{High Institutional Credibility}  \\ \hline
    \textbf{Low Political Polarization} & Negligible anticipated impacts. Traditional Professionalism is Expected by the Military, Civilian Leaders, and the American Public & Possible Impacts. There is a demand for military voice, but little substantive matter stemming from US domestic politics that can cause crucial disagreement. Little incentive to deviate from traditional concepts of loyalty and duty. \\ \hline
    \textbf{High Political Polarization} & Possible Impacts. Polarization raises the stakes of domestic politics, but military professionals are constrained from voicing their opinions and/or interfacing with the public due to low institutional credibility with the American public &  Highly Likely Impacts to Professionalism. Likely impacts. Military professionals have a "platform with the public" due to high institutional credibility, and high polarization raises the likelihood of disagreement between individual military elites and civilian leaders on a number of crucial topics.\\ \hline
\end{tabular}
\label{table:comparison} % is used to refer this table in the text
\end{footnotesize}
\end{table}
\doublespacing
The second possible set of conditions is when political polarization is low and institutional credibility is high. When these set of conditions exist, the military has a platform to interact with the public, as the public highly respects the military institution. The military enjoys this high level of trust and does not want to lose it. Further, because the degree of political polarization is low, there is little incentive for the military as an institution or military elites to use the trust they enjoy with the American public to adopt or denounce an issue associated with any partisan entity. The high institutional credibility increases the chance of the military interacting with the public in an unprofessional way, but the low political polarization removes the incentives of doing so.

@@

The final set of conditions exists when political polarization and institutional credibility are high. It is in this world that I contend the American military finds itself in now in 2020. Under these conditions, the military faces dual dangers. High institutional credibility risks the military institution unilaterally acting or another body co-opting the military institution to speak out an a number of issue areas, and simultaneously, the highly polarized environment risks the military advocating for or denouncing partisan sentiments in the name of upholding the very traits it is most adhered to - loyalty and duty. Indeed, this set of conditions comprise a very dangerous world for the military institution and the military professional.

# Methodological Approach, Data, and Hypotheses

I now aim to see whether the discourse of retired military officers has changed over time, specifically since the 1970s, as polarization has worsened and as the institutional credibility of the military has increased. To answer this question, I assemble and examine a data set that consists of opinion pieces authored by retired military officers in major US newspapers. My methodological approach involves two key steps. First, I determine if there have been shifts in the topics that retired military officers have been writing about. Second, I apply a normative lens to determine if there have been shifts in the tone and tenor of the pieces authored by these officers, particularly regarding the professional aspects of the military previously discussed. In simple terms, I want to answer the questions, what are these retired military officers writing about, and through the lens of professionalism, should they be saying what they are saying?

Using the Factiva and Nexus-Uni databases, I conduct a search that focuses keywords in the byline (author field) such as "retired military officer," "retired from the Army, Navy, Marine Corps, Air Force," "former Chairman of the Joint Chiefs of Staff," "General," "Admiral," and so forth. For sources, I confine my search to the largest US newspapers that regularly run opinion-editorial pieces, which includes  _The Wall Street Journal_ (WSJ), _The New York Times_ (NYT), _The Washington Post_ (WaPo), _The Los Angeles Times_, and _The USA Today_. I begin the search for the data of January 1, 1979 - 30 June 2020^[I also open my search to include both print and online offerings to ensure that I am capturing any opinion pieces published digitally. Most of the newspapers initiated digital offerings in the late 1990s.]. Using newspapers, and particularly these specific ones, provides some advantages as well as disadvantages. The primary advantage is that these sources have been around since the period I wish to examine, unlike many of the online only sites that have only been present in the digital age. This allows me to form a data set which covers more time. The primary disadvantage is that I am not capturing all of the opinion-pieces authored or given by retired military officers. For example, retired military officers are common on cable TV, often weighing in on topics, and I am not capturing those here. It is also true that over time,  social media has taken over as the primary means in which a majority of Americans receive their news [@martin_how_2018]. Print newspapers still exist, but overall circulation is still down today compared to earlier years, even after factoring in the expanse of digital options available^[As a means of rough comparison, estimated total daily newspaper circulation in the US in 2000 was 62 million and 54 million for weekday and weekend editions, respectively, while in 2018, the the same figures were 29 million and 31 million [@barthel_leading_2019]]. Nonetheless, I contend that the data I am able to assemble will still be informative. Publishing opinion-editorial pieces, especially in major, competitive newspapers, is a highly competitive process. For the newspapers that I examine in this study, it is not uncommon to see opinion pieces published by figures that include heads of state, members of congress, governmental figures, and leaders of industry and academia. Additionally, opinion pieces are deliberately authored for the purpose of informing and persuading the public on a salient or timely topic. They are different from interviews or commentary that may be asked by a questioner because the author must take time to put forth his or her position in writing. The data therefore allows the researcher to truly ascertain what the specific views and arguments of an author were on a given topic or policy.

Of note, when assembling the data, some authors are explicitly identified in some of their publications as retired military officers, but not in others.^[This was true of several individuals, including Alexander Haig, a retired Army General who later served as Secretary of State under Ronald Reagan;  Robert McFarlane, a retired Marine Corps Lieutenant Colonel who later served as the National Security Adviser to President Reagan, Andrew Bacevich, a retired Army Colonel who then became a prominent historian and author; and Michael Hayden, a retired Air Force General who served as director of the National Security Agency and the Central Intelligence Agency]. I conduct a separate search for all opinion pieces penned by these authors, regardless of whether they were identified as a retired military officer or not, concluding that it is best do so for coding purposes. Ultimately, the data set contains 334 opinion-editorial publications, penned by 178 different authors or author teams.^[There are several co-authored pieces. For example, where General (Retired) David Petraeus wrote one publication and another one was written by General (Retired) David Petraeus and Michael O'Hanlon, I counted these as two different authors for coding purposes.] 10 of these pieces are written unanimously, with nods by the authors to their previous service.

From a macro perspective, consider for a moment all that the date range includes with respect to the US military. Indeed, numerous and substantive events, debates, and changes have occurred in this time period for retired military officers to weigh in on. In terms of external threats and wars, this period includes the final years of the Cold War against the Soviet Union, the first war in Iraq, operations in the Balkans, and the post-9/11 wars in Iraq and Afghanistan, as well as a host of smaller but important military operations in Grenada, Panama, Somalia, and Haiti. The military has also undergone significant social and institutional changes in the time period examined. In the late 1970s, the military had ended the draft only years before following the conclusion of the Vietnam War. The debates over "Don't Ask, Don't Tell" and policies regarding women in the combat arms are also included in the data, as are numerous discussions of military budgets, weapons systems, and alliances. 
Attempting to reconcile the fact that the military as an institution prides itself on being a non-partisan and apolitical institution, that newspaper circulations (even after accounting for digital offerings) are substantially down over the past 40 years, and that the US political environment is increasingly polarized along partisan lines, I adopt the following hypotheses:

@@

_$H_2$: Because overall newspaper consumption has dropped in the past 40 years, there will be fewer opinion pieces over time authored by military officers._

_$H_3$: Given that polarization has become increasingly pronounced in the United States in recent years, there will be an increase in the number of op-eds which reveal political and partisan sentiments (indication of a drop in professionalism)._

# First Layer Analysis: Topic Addressed

In the first level of analysis, after capturing key biographical and publication information, such as author name, article title, source, publication date, and branch of service and rank of the author, I code each observation into one of several binary categories that corresponds to the overall subject matter of the publication. The goal is to create a coding scheme by subject that is sufficiently broad but also specific enough that is helpful in differentiating the overall main point of each publication. The following coding scheme is therefore devised and implemented (all variables are binary, taking on the value of 1 if the publication corresponds to the general definition of the variable, and 0 otherwise): \singlespace

* **War_Ops**- if the piece advocates for or denounces a particular course of action directly related to operational or war fighting decisions of a clear military nature. Examples include: the wisdom of troop increases, the Surge in Iraq, whether to bomb Serbia, or even the role of the prison at Guantanamo.

* **Trp_Clt**- if the piece generally showcases or highlights the work and sacrifice of members of the armed forces; to defend the work of specific members of the Armed Forces; or to call for policing up the institutional cultures of the services. Examples include: remembering our fallen soldiers on Memorial Day; improving the institutional culture of the Army or Navy; advocating on behalf of or remembering a specific flag officer.

* **Str_Adv**- if the piece is generally written to offer some sort of strategic advice for the nation, to include the wisdom of pursuing particular alliances or agreements, or adopting a specific weapons platform or military budget. Examples include: why the US needs to strengthen NATO, build ties with China, increase defense spending, or not retire the A-10 combat plane.

* **MilPSoc**- if the piece advocates for or denounces a particular social policy that _directly influences_ the armed forces. Examples include: debates over keeping or removing the "Don't Ask, Don't Tell" policy, opening up combat roles to women, the role of Transgender members of the armed forces.

* **Dom_P**- if the piece advocates for or denounces a position on a policy that _does not directly_ influence the armed forces in its day to day activities. Examples include: support for the second amendment, immigration policy, and whether confederate statues should be taken down or left up.

* **Fig_Name**- if the piece advocates for or denounces a particular position held by a political administration, or _if the main point of the piece_ is to explicitly defend or attack a political figure by name. Examples: endorsing a Presidential candidate for office, or for attacking a President or member of Congress for a decision that was made.

\doublespace
Separating opinion-editorial publications into one of several distinct categories is an imperfect process. To aid the coding process, I first identify the thesis statement of each opinion piece, which in almost every case, makes possible a reasonably clear classification of the publication. Two points are worth mentioning. First, I attempt to keep in mind the fact that opinion pieces are intended to argue for something, and thus are, to an extent, designed to be provocative. With this in mind, I did not automatically code pieces that generally advocated for or cautioned the President in the **Admin_Fig_Name** category. Second, there were some instances in which I had to choose between placing a publication into one of two categories, say, between **Warfighting_Ops** and **Strat_Advice**, but very few if any instances in which I had to choose between categories that differed greatly, such as **TroopSptCulture** and **Admin_Fig_Name**. Table 1 reveals the breakdown of the data by article type.

@@

```{r table2, fig.align="center", include=TRUE, echo=FALSE}
countbytype %>%
  kable(format="latex", caption="Publications by Subject") %>%
  kable_styling(latex_options = "striped", font_size=10, full_width = T) %>% 
  row_spec(0, bold=T)
```

Table 2 reveals that an overwhelming majority (289 of 334, or 86.5%) of the total number of opinion pieces written since 1979 in the major newspapers examined can be classified according to one of the following three categories: **War_Ops**, **Trp_Clt**, and **Str_Adv**. It is reasonable to conclude that these op-eds written are appropriate from the perspective of civil-military relations in that they address topics which American citizens can reasonably expect retired military officers to bring their expertise to bear upon. The remaining 45 opinion pieces (13.5% of the total) fall into one of the other three categories: **MilPSoc**, **Dom_P**, and **Fig_Nme** and require further examination.  They are not necessarily inappropriate, but they likely address topics that may be beyond reasonable military expertise or which reveal partisan preferences. This second layer of analysis will help expose whether there is any indication that professionalism as a norm among retired military officers appears to be weakening.

>>>>

# When Words Become Weaponized - A Theory of Political Polarization and Military Professionalism

Figure One, below, shows describes a theory for how the variables of political polarization, institutional credibility, military professionalism, and other military outcomes may relate to each other. The theory contends that increasing political polarization ultimately reduces military professionalism, but that there may also be an intervening variable, rising institutional credibility, that also is impacted by political polarization and yet further impacts military professionalism. Finally, though it is outside the scope of this paper, I contend that an increase in political polarization may at some point cause discernible changes in various military outcomes. Such outcomes may include a multitude of topics, but would include items such as military effectiveness, to include the military's ability to fight and win battles [@talmadge_dictators_2015], and to formulate sound strategic advice [@rapp_civil-military_2015]. Indeed, the full degree to which and in what ways polarization will impact the military are not yet fully known.

@@

problem_dag %>% tidy_dagitty() %>% 
  dag_label(labels=c("MO"="Changes in\n Military\n Outcomes", 
                     "MP"="Reduction in\n Military\n Professionalism", 
                     "PP"="Increasing\n Political\n Polarization", 
                     "IC"="Rising\n Institutional\n Credibility")) %>% 
  ggplot(aes(x=x, y=y, xend=xend, yend=yend))+
  geom_dag_edges()+
  geom_dag_point()+
  geom_dag_text()+ 
  geom_dag_label_repel(aes(label=label, fill=label), col="white", show.legend=FALSE)+ theme_dag() + labs(title="Figure One", subtitle="The Impact of Political Polarization on Military Professionalism")

@@

What mechanisms might be at work through increasing political polarization that ultimately impact military professionalism? To answer this important question, I rely on the three major characteristics of the contemporary polarized political landscape previously identified by scholars of American Politics. These included, first, a sharp increase in the number and scope of issues that have been subsumed by partisan debate and disagreement (the phenomenon referred to as "conflict extension" [@layman_party_2002]); second, the increase in the strength of negative feelings held by one partisan group towards those of the other side (the rise of "negative partisanship" [@abramowitz_great_2018]); and third, an increase in the salience of debates over issues that individual partisans consider core to their individual identities [@white_values_2002; @mann_its_2012; @sides_identity_2018]. Each of these three features of the politically polarized landscape strain the concept of military professionalism in important ways, and taken together, comprise a significant challenge to the individual military professional and thus, to the professional nature of the military as a whole. These challenges are summarized in Table One, below.

\singlespacing
\begin{table}[ht]
\begin{footnotesize}
\caption{Mechanisms of Rising Political Polarization Which Impact Military Professionalism} % title of Table
\centering % used for centering table
\begin{tabular}{|p{5cm}|p{5cm}|p{5cm}|p{5cm}|}
    \hline
    \textbf{Characteristics of Increasing Political Polarization} & \textbf{Description} & \textbf{Resulting Strain on Military Professionalism}  \\ \hline
    \textbf{Conflict Extension} & Less politically-neutral space exists as partisan disagreements subsume a greater number of issues & Military Professionals face an increasingly difficult prospect of remaining apolitical. \\ \hline
    \textbf{Negative Partisanship} & Strong feelings of personal dislike and disdain for those who hold to opposing political views and who are on the "other side" & Greater willingness to confront or take issue with those with opposing views\\ \hline
    \textbf{Core and dearly-held issues are increasingly at stake} & The salience of one's identity and belonging to a particular group is raised; Belief that moral right and wrong are at stake & Changes in one's concept of loyalty \\ \hline
\end{tabular}
\label{table:comparison} % is used to refer this table in the text
\end{footnotesize}
\end{table}
\doublespacing

The first way in which the concept of military professionalism is challenged as a result of increasing political polarization is that there is increasing difficulty for individuals to remain politically neutral, non-partisan, or otherwise apolitical - a hallmark of military professionalism. This occurs because as "conflict extension" [@layman_party_2002] occurs and an increasing number of issues are swallowed up by partisan battlegrounds, there is less politically neutral space for someone to stand on. When polarization levels in America were not as high as they are today, it was more possible for individuals to either stay out of politics, avoid partisan squabbles, and otherwise to stay politically neutral. Today's polarizing environment seems to have the opposite impact: polarized politics seem to thrust issues and the unfolding political fight at the individual citizen, even when the individual is not deliberately seeking confrontation. There is no theoretical reason to suspect that this mechanism would not impact military leaders. Military leaders are citizens of the Republic, and though the institutional ethos of the military profession is to remain apolitical, such an ideal is difficult to uphold if less and less neutral political space exists.

A second challenge to military professionalism stems from the "negative partisanship"[@abramowitz_great_2018] that dominates the polarized American political landscape. Strong feelings towards those on a host of other issues are becoming increasingly hostile. These feelings are growing increasingly strong likely because of the other prominent features of political polarization: a decrease in the availability of neutral political space, and the importance of issues that are at stake in many of today's political debates.

A third, and I contend a more insidious, mechanism through which increasing political polarization works to impact military professionalism is by altering one's sense of loyalty and identity. This occurs because the issues that are at stake as political polarization increases are in many instances involving issues that individuals associate with core concepts of one's identity, and thus, one's sense of moral right and wrong. Because individuals care about their identities and what they believe to be right and wrong, they are willing to defend these values, even at a cost. The rise of "identity politics" in America as political polarization has worsened illustrates this point.

However, it is also precisely on this point that the military institution is perhaps most vulnerable to increasing political polarization. I contend that the rise of identity politics can and will be harmful if a military professional's identity and the loyalty associated with that identity directly conflicts with his or her identity and loyalty as a professional soldier. As any soldier knows, and as civilians can reasonably understand,few traits are more sacred in the military than loyalty. Indeed, the virtue of loyalty is extolled in a host of codes and mantras that have played a central part of the American military since its inception, and loyalty is captured, either in word or in spirit, in a host of phrases and professional codes. In the middle of the the West Point Cadet Prayer is the eloquently written plea, "make us to choose the harder right instead of the easier wrong" [@wheat_west_1918].

High levels of political polarization obfuscate the ideal of loyalty, however. As core issues that involve individual perceptions of moral right and wrong are at stake, individuals, to include military officers, are less likely to not become involved, attached, or otherwise invested in these issues. This is especially true when the language and appeal of concepts of loyalty are employed to justify one is or is not taking action over an issue. In a highly polarized political climate, there is increasing likelihood that two different people military officers may adopt completely different opinions regarding a contentious issue, each believing that the other is deeply and perhaps even morally wrong.

@@

# Accounting for Institutional Credibility

The previous section described how increasing political polarization presents significant challenges to the concept of military professionalism. But when will these challenges become public knowledge, and perhaps more relevant, how does increasing political polarization impact the credibility of the military institution?

One can imagine several possible scenarios. It is reasonable to consider that as political polarization increases, respect and trust in seemingly partisan institutions would decrease, while trust in nonpartisan and apolitical institutions would rise. But why? Would such a trend reflect a genuine increase in trust in the nonpartisan institution, or would it reflect the simple consequence that as one institution becomes less respected, others might simply gain as a result? Certainly, more work needs to be done in this area.

Nonetheless, I contend that when institutional credibility is high, a demand signal is created by the American public such that future interactions between the military institution and the American public are encouraged and welcomed. From an institutional perspective, the military might respond to the public's demand signal for increased military interaction by increasingly participating in public ceremonies, parades, and other showcases of military prowess (at sporting venues, for instance). From an individual perspective, military elites, especially retired military officers, would be psychologically primed to increasingly dialogue with the American public when military credibility is high. By extension, other outlets such as the press would also be primed to showcase military elites when military credibility runs high, because these elites represent an institution that commands a high degree of public respect.

Ultimately, four possible worlds exist when the impacts of military credibility and political polarization are considered in tandem (see Table Two, below). The first is when political polarization and military credibility are both low. When these conditions hold, there is no anticipated drop in military professionalism by the military institution nor its individual elites. With these conditions, the military does not have an out-sized platform to interact with the American public, nor are there outstanding incentives for the military to raise significant disagreement over any domestic political issues.

\singlespacing
\begin{table}[ht]
\begin{footnotesize}
\caption{Table Two: Four Types of "Military Professionalism Worlds" Where Institutional Credibility and Political Polarization Vary} % title of Table
\centering % used for centering table
\begin{tabular}{|p{5cm}|p{5cm}|p{5cm}|p{5cm}|}
    \hline
    Influence of Variables & \textbf{Low Institutional Credibility} & \textbf{High Institutional Credibility}  \\ \hline
    \textbf{Low Political Polarization} & Negligible anticipated impacts. Traditional Professionalism is Expected by the Military, Civilian Leaders, and the American Public & Possible Impacts. There is a demand for military voice, but little substantive matter stemming from US domestic politics that can cause crucial disagreement. Little incentive to deviate from traditional concepts of loyalty and duty. \\ \hline
    \textbf{High Political Polarization} & Possible Impacts. Polarization raises the stakes of domestic politics, but military professionals are constrained from voicing their opinions and/or interfacing with the public due to low institutional credibility with the American public &  Highly Likely Impacts to Professionalism. Likely impacts. Military professionals have a "platform with the public" due to high institutional credibility, and high polarization raises the likelihood of disagreement between individual military elites and civilian leaders on a number of crucial topics.\\ \hline
\end{tabular}
\label{table:comparison} % is used to refer this table in the text
\end{footnotesize}
\end{table}
\doublespacing
The second possible set of conditions is when political polarization is low and institutional credibility is high. When these set of conditions exist, the military has a platform to interact with the public, as the public highly respects the military institution. The military enjoys this high level of trust and does not want to lose it. Further, because the degree of political polarization is low, there is little incentive for the military as an institution or military elites to use the trust they enjoy with the American public to adopt or denounce an issue associated with any partisan entity. The high institutional credibility increases the chance of the military interacting with the public in an unprofessional way, but the low political polarization removes the incentives of doing so.

@@

The final set of conditions exists when political polarization and institutional credibility are high. It is in this world that I contend the American military finds itself in now in 2020. Under these conditions, the military faces dual dangers. High institutional credibility risks the military institution unilaterally acting or another body co-opting the military institution to speak out an a number of issue areas, and simultaneously, the highly polarized environment risks the military advocating for or denouncing partisan sentiments in the name of upholding the very traits it is most adhered to - loyalty and duty. Indeed, this set of conditions comprise a very dangerous world for the military institution and the military professional. 
# Methodological Approach, Data, and Hypotheses

I now aim to see whether the discourse of retired military officers has changed over time, specifically since the 1970s, as polarization has worsened dramatically. To answer this question, I assemble and examine a data set that consists of opinion-editorial publications authored by retired military officers in major US newspapers. My methodological approach involves two fundamental steps. First, I determine what topics retired military officers have been writing about over time. Second, I apply a normative lens to determine if there have been shifts in the tone and tenor of the pieces authored by these officers, particularly regarding the professional aspects of the military previously discussed (apolitical in nature, non-partisan, relying on relevant military expertise, etc.). In essense, I want to answer the questions, what are these retired military officers writing about, and through the lens of professionalism, should they be saying what they are saying? Is there any indication that increasing political polarization is altering what and how these retired officers are communicating with the American public?

To assemble the data, I use the Factiva and Nexus-Uni databases, conducting a search that focuses keywords in the byline (author field) such as "retired military officer," "retired from the Army, Navy, Marine Corps, Air Force," "former Chairman of the Joint Chiefs of Staff," "General," "Admiral," and so forth. I confine my search to the largest US newspapers that regularly run opinion-editorial pieces, which includes  _The Wall Street Journal_ (WSJ), _The New York Times_ (NYT), _The Washington Post_ (WaPo), _The Los Angeles Times_, and _The USA Today_. The time period for the data includes the period January 1, 1979 - 30 June 2020, the main limitation being that not all of the Newspaper sources are available for previous dates. I also open my search to include both print and online offerings to ensure that I am capturing any opinion pieces published digitally.

The sources that I rely upon provide some relative advantages, but they are not without limitations. The primary advantage is that these sources have some longevity; they are not modern blogging sites or online-only offerings. As a result, I can truly examine a larger period of time and thus, examine whether how political polarization has impacted the way in which military officers engage with the American Public. Simultaneously, though, this presents a disadvantage. Retired military officers engage with the American public through a host of mediums, to include social media, blogging sites and cable television, neither of which I am capturing in my data set. It is also true that over time, social media has taken over as the primary means in which a majority of Americans receive their news [@martin_how_2018]. Print newspapers still exist, but overall circulation is still down today compared to earlier years, even after factoring in the expanse of digital options available^[As a means of rough comparison, estimated total daily newspaper circulation in the US in 2000 was 62 million and 54 million for weekday and weekend editions, respectively, while in 2018, the the same figures were 29 million and 31 million [@barthel_leading_2019]]. Nonetheless, I contend that the data I am able to assemble will still be informative. Publishing opinion-editorial pieces, especially in major, competitive newspapers, is a highly competitive process. For the newspapers that I examine in this study, it is not uncommon to see opinion pieces published by figures that include heads of state, members of congress, governmental figures, and leaders of industry and academia. Additionally, opinion pieces are deliberately authored for the purpose of informing and persuading the public on a salient or timely topic. They are different from interviews or commentary that may be asked by a questioner because the author must take time to put forth his or her position in writing. The data therefore allows the researcher to truly ascertain what the specific views and arguments of an author were on a given topic or policy.

Of note, when assembling the data, some authors are explicitly identified in some of their publications as retired military officers, but not in others.^[This was true of several individuals, including Alexander Haig, a retired Army General who later served as Secretary of State under Ronald Reagan;  Robert McFarlane, a retired Marine Corps Lieutenant Colonel who later served as the National Security Adviser to President Reagan, Andrew Bacevich, a retired Army Colonel who then became a prominent historian and author; and Michael Hayden, a retired Air Force General who served as director of the National Security Agency and the Central Intelligence Agency]. I conduct a separate search for all opinion pieces penned by these authors, regardless of whether they were identified as a retired military officer or not, concluding that it is best do so for coding purposes.

Ultimately, the data set contains 334 opinion-editorial publications, penned by 178 different authors or author teams.^[There are several co-authored pieces. For example, where General (Retired) David Petraeus wrote one publication and another one was written by General (Retired) David Petraeus and Michael O'Hanlon, I counted these as two different authors for coding purposes.] 10 of these pieces are written unanimously, with nods by the authors to their previous service. From a macro perspective, consider for a moment all that the date range includes with respect to the US military. Indeed, numerous and substantive events, debates, and changes have occurred in this time period for retired military officers to weigh in on. In terms of external threats and wars, this period includes the final years of the Cold War against the Soviet Union, the first war in Iraq, operations in the Balkans, and the post-9/11 wars in Iraq and Afghanistan, as well as a host of smaller but important military operations in Grenada, Panama, Somalia, and Haiti. The military has also undergone significant social and institutional changes in the time period examined. In the late 1970s, the military had ended the draft only years before following the conclusion of the Vietnam War. The debates over "Don't Ask, Don't Tell" and policies regarding women in the combat arms are also included in the data, as are numerous discussions of military budgets, weapons systems, and alliances.

Attempting to reconcile the fact that the military as an institution prides itself on being a non-partisan and apolitical institution, that newspaper circulations (even after accounting for digital offerings) are substantially down over the past 40 years, and that the US political environment is increasingly polarized,  I adopt the following hypotheses:

@@

_$H_2$: Because overall newspaper consumption has dropped in the past 40 years, there will be fewer opinion pieces over time authored by retired military officers._

_$H_3$: Given that polarization has become increasingly pronounced in the United States over the time period examined, there will be an increase in recent years in the number of op-eds which reveal political and partisan sentiments, indicating a drop in military professionalism._

# First Layer Analysis: Topics Addressed by Retired Military Officers

In the first level of analysis, I code each observation into one of several binary categories that corresponds to the overall subject of the publication. The goal is to create a coding scheme that captures a concise number of topical descriptions that are still sufficiently broad enough to reveal variation in the topic addressed by retired military authors. I develop a total of six topical variables, and each observation takes on the value of 1 if the publication corresponds to the general definition of the variable, and 0 otherwise.

These topical variables are: warfighting and operations, if the piece advocates for or denounces a particular course of action directly related to operational or war fighting decisions of a clear military nature; troop support and institutional culture, if the publication generally highlights the work and sacrifice of members of the armed forces, defends the work of specific members of the Armed Forces, or makes a call for action to improve a service's institutional culture; strategic advice, if the piece is generally written to offer some sort of strategic advice for the nation, to include the wisdom of pursuing particular alliances or agreements, or adopting a specific weapons platform or military budget; social policy directly related to the military, if the piece advocates for or denounces a particular social policy that _directly influences_ the armed forces (Examples include: debates over keeping or removing the "Don't Ask, Don't Tell" policy, opening up combat roles to women, the role of Transgender members of the armed forces); domestic policy, if the piece advocates for or denounces a position on a policy that _does not directly_ influence the armed forces in its day to day activities (Examples include: support for the second amendment, immigration policy, and whether confederate statues should be taken down or left up); figures by name, if the piece advocates for or denounces a particular position held by a political administration, or _if the main point of the piece_ is to explicitly defend or attack a political figure by name (Examples: endorsing a Presidential candidate for office, or for attacking a President or member of Congress for a decision that was made).

Separating opinion-editorial publications into one of several distinct categories is an imperfect process, and requires as objective a reading of the publication as possible. To aid the coding process, I identify the thesis statement of each opinion piece, which in most cases is identifiable in either the lead or concluding paragraph of the opinion piece. In most cases, identifying and re-reading the thesis of each op-ed makes it possible to reasonably classify the publication by topic. Two additional points, however, are worth emphasizing here.

First, opinion pieces are intended to argue for something, and thus are, to an extent, designed to be at least mildy provocative. With this in mind, I did not necessarily code pieces that called for a political figure or instituton to take a specific action into the category of _figures by name_.  Again, the identification of the publication's thesis statement aided significantly in identifying the main thrust of the argument, and thus, in classifying the argument by topic. Second, there were some instances in which I had to choose between placing a publication into one of two categories. However, in instances of choosing between two categories, the choice to classify between categories was usually between categories that were reasonably close. For example, there were a few instances of choosing to classify a publication between the warighting and operations category and the strategic advice category, but no instances of choosing between the warfighting and operations category and the domestic policy category. Table 3 reveals the breakdown of all examined publications by article type.

@@

```{r table3, fig.align="center", include=TRUE, echo=FALSE}
countbytype %>%
  kable(format="latex", caption="Publications by Subject") %>%
  kable_styling(latex_options = "striped", font_size=10, full_width = T) %>% 
  row_spec(0, bold=T)
```

Table 3 reveals that an overwhelming majority (289 of 334, or 86.5%) of the total number of opinion pieces written since 1979 in the major newspapers examined can be classified according to one of the following three categories: **War_Ops**, **Trp_Clt**, and **Str_Adv**. It is reasonable to conclude that these op-eds written are appropriate from the perspective of civil-military relations in that they address topics which American citizens can reasonably expect retired military officers to bring their expertise to bear upon. The remaining 45 opinion pieces (13.5% of the total) fall into one of the other three categories: **MilPSoc**, **Dom_P**, and **Fig_Nme** and require further examination.  They are not necessarily inappropriate, but they likely address topics that may be beyond reasonable military expertise or which reveal partisan preferences. This second layer of analysis will help expose whether there is any indication that professionalism as a norm among retired military officers appears to be weakening.
